---
title: "Tree sizes"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: 
  gfm:
    toc: true
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

set.seed(123)
ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

Are the planted trees reaching the canopy yet?

## DBH

Looking at DBH recorded in the latest full census of the SBE.

```{r}
#| output: FALSE

library("tidyverse")
library("here")
library("patchwork")
```

```{r}
new_census <- 
  readRDS(here::here("data", "derived", "data_cleaned.rds")) %>% 
  filter(census_id == "full_measurement_03") 

new_census %>% 
  drop_na(dbh_mm) %>% 
  glimpse()
```

```{r}
new_census %>% 
  summarise(
    mean = mean(dbh_mm, na.rm = TRUE),
    median = median(dbh_mm, na.rm = TRUE),
    max = max(dbh_mm, na.rm = TRUE),
    min = min(dbh_mm, na.rm = TRUE)
  )
```

```{r}
new_census %>% 
  ggplot(aes(x = dbh_mm)) +
  geom_density() +
  labs(x = "DBH /mm")
```


```{r}
new_census %>% 
  ggplot(aes(x = dbh_mm)) +
  geom_density() +
  labs(x = "DBH /mm") +
  facet_wrap(~treatment)
```

### Trees above 30cm DBH

```{r}
new_census %>% 
  filter(dbh_mm > 300) %>% 
  nrow()
```

```{r}
new_census %>% 
  filter(dbh_mm > 300) %>% 
  group_by(genus_species) %>% 
  summarise(n_distinct(plant_id)) %>% 
  arrange(`n_distinct(plant_id)`
          |> desc() ) %>% 
  knitr::kable()
```

```{r}
new_census %>% 
  filter(dbh_mm > 300) %>% 
  group_by(treatment) %>% 
  summarise(n_distinct(plant_id)) %>% 
  arrange(`n_distinct(plant_id)`
          |> desc() ) %>% 
  knitr::kable()
```

## Height

Using a Weibull function to estimate tree height from DBH:

predicted height = _a_\*(1 − exp(−_b_\*DBH^_c_^))

Parameters _a, b_ and _c_ from [Feldpausch _et al._ 2012](https://doi.org/10.5194/bg-9-3381-2012) (Asia-specific model).

_We think they use DBH in cm and height in m_

```{r}
a <- 57.122	

b <- 0.0332	

c <- 0.8468	
```

```{r}
get_height <- function(dbh, a, b, c) {
  height <-
    a * (1-exp(-b * (dbh)^c))
  
  return(height)
}
```

So for a tree with DBH of 50 cm, the predicted height is:

```{r}
get_height(dbh = 50,
           a = a,
           b = b,
           c = c)
```

Across the all the SBE planted trees:

```{r}
new_census %>% 
  mutate(dbh_cm = dbh_mm / 10) %>% 
  mutate(tree_height_m = get_height(dbh = dbh_cm,
                                     a = a,
                                     b = b,
                                     c = c) ) %>% 
  summarise(
    mean = mean(tree_height_m, na.rm = TRUE),
    median = median(tree_height_m, na.rm = TRUE),
    max = max(tree_height_m, na.rm = TRUE),
    min = min(tree_height_m, na.rm = TRUE)
  )
```

```{r}
new_census %>% 
  mutate(dbh_cm = dbh_mm / 10) %>% 
  mutate(tree_height_m = get_height(dbh = dbh_cm,
                                     a = a,
                                     b = b,
                                     c = c) ) %>% 
  ggplot(aes(x = tree_height_m)) +
  geom_density() 
```

```{r}
new_census %>% 
  mutate(dbh_cm = dbh_mm / 10) %>% 
  mutate(tree_height_m = get_height(dbh = dbh_cm,
                                     a = a,
                                     b = b,
                                     c = c) ) %>% 
  ggplot(aes(x = tree_height_m)) +
  geom_histogram() 
```

By treatment:

```{r}
new_census %>% 
  mutate(dbh_cm = dbh_mm / 10) %>% 
  mutate(tree_height_m = get_height(dbh = dbh_cm,
                                     a = a,
                                     b = b,
                                     c = c) ) %>% 
  ggplot(aes(x = tree_height_m)) +
  geom_density() +
  facet_wrap(~treatment)
```

```{r}
new_census %>% 
  mutate(dbh_cm = dbh_mm / 10) %>% 
  mutate(tree_height_m = get_height(dbh = dbh_cm,
                                     a = a,
                                     b = b,
                                     c = c) ) %>% 
  ggplot(aes(x = tree_height_m)) +
  geom_histogram() +
  facet_wrap(~treatment)
```

```{r}
new_census %>% 
  mutate(dbh_cm = dbh_mm / 10) %>% 
  mutate(tree_height_m = get_height(dbh = dbh_cm,
                                     a = a,
                                     b = b,
                                     c = c) ) %>% 
  ggplot(aes(x = tree_height_m)) +
  geom_histogram() +
  facet_wrap(~treatment)
```

By species:

```{r}
new_census %>% 
  mutate(dbh_cm = dbh_mm / 10) %>% 
  mutate(tree_height_m = get_height(dbh = dbh_cm,
                                     a = a,
                                     b = b,
                                     c = c) ) %>% 
  drop_na(genus_species) %>% 
  ggplot(aes(x = tree_height_m)) +
  geom_density() +
  facet_wrap(~genus_species)
```

```{r}
new_census %>% 
  mutate(dbh_cm = dbh_mm / 10) %>% 
  mutate(tree_height_m = get_height(dbh = dbh_cm,
                                     a = a,
                                     b = b,
                                     c = c) ) %>% 
  drop_na(genus_species) %>% 
  ggplot(aes(x = tree_height_m)) +
  geom_histogram() +
  facet_wrap(~genus_species)
```

