---
title: "Mapping the SBE seedlings"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: 
  gfm:
    toc: true
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

set.seed(20440)
ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

```{r}
#| output: FALSE

library("tidyverse")
library("here")
library("patchwork")
```

> Andy explained the plan to compare seedling height with local canopy height from LiDAR, which I think is a great idea! It would be great if we could do this at a spatial scale finer than the 4-ha plots, because there are lots of gaps and spaces in between trees where the seedlings may grow faster / differently.  Can you tell which planting line the seedlings are in, and roughly how far along it they are? Do you think we could work at a 10x10 m or 20x20 m scale for this analysis?

We know the planting line for each seedling and it's position along the line.
There are 3 m between each planting position and 10 m between each line. 

```{r}
data <- 
  readRDS(here::here("data", "derived", "data_cleaned.rds")) %>% 
  filter(census_id == "full_measurement_03")
```

```{r}
data %>% 
  filter(plot %in% c("001", "002", "003", "004", "005", "006")) %>% 
  ggplot(aes(x = line, y = position, colour = survival)) +
  geom_point(size = 0.25) +
  facet_wrap(~plot)
  
```

```{r}
#| fig-height: 5
#| fig-width: 6
data %>% 
  filter(plot == "001") %>% 
  ggplot(aes(x = line, y = position, colour = survival)) +
  geom_point(size = 0.5) +
  ggtitle("plot 1")
  
```
