---
title: "First models - SBE 2nd decade"
---

```{r setup}
#| include: false

document_name <- tools::file_path_sans_ext(
	basename(knitr::current_input())
)

knitr::opts_chunk$set(
	fig.path = file.path("figures", document_name, "")
)

set.seed(20689)

source(here::here("code", "functions", "plotting.R"))
ggplot2::theme_set(theme_sbe())
```

First step to fit:

`partitioning_component ~ treatment + block + (1 | species_mix)`

Later we will fit models for the 4-species plots only:

`partitioning_component ~ structural_complexity * generic_richness + block + (1 | species_mix)`

The 4-species analysis is treated as a separate subanalysis because structural complexity and generic richness are only defined within that subset.

```{r}
library("tidyverse")
library("patchwork")
library("lme4")
library("glmmTMB")
library("broom.mixed")
library("DHARMa")
```

```{r}
partitions <- readRDS(here::here("data", "derived", "biodiv_effects.rds"))
```

Create model fitting function that we can run over multiple `partitioning_component`s

```{r}
fit_models <- function(data, formula, method = c("lmer", "lm", "glmmTMB")) {
	responses <- c(
		m_NE = "net",
		m_CE = "compl",
		m_CE_size = "size_compl",
		m_CE_dens = "dens_compl",
		m_SE = "selec",
		m_SE_size = "size_selec",
		m_SE_dens = "dens_selec"
	)

	fit_one <- purrr::quietly(\(response) {
		if (method == "lmer") {
			lme4::lmer(
				stats::as.formula(
					paste(response, formula)
				),
				data = data
			)
		} else if (method == "glmmTMB") {
			glmmTMB::glmmTMB(
				stats::as.formula(
					paste(response, formula)
				),
				data = data,
				family = t_family(link = "identity")
			)
		} else {
			stats::lm(
				stats::as.formula(
					paste(response, formula)
				),
				data = data
			)
		}
	})

	results <- purrr::map(responses, fit_one)

	# check for warnings and singular fits
	tibble::tibble(
		name = names(responses),
		fit = purrr::map(results, "result"),
		singular = purrr::map_lgl(
			results,
			\(x) method == "lmer" && lme4::isSingular(x$result)
		),
		warning = purrr::map_chr(
			results,
			\(x) {
				warnings <- x$warnings

				if (method == "lmer" && lme4::isSingular(x$result)) {
					warnings <- c(
						warnings,
						"boundary (singular) fit: see help('isSingular')"
					)
				}

				paste(unique(warnings), collapse = "\n")
			}
		)
	)
}
```

## `lmer` models

```{r}
models_lmer <- fit_models(
	data = partitions,
	formula = "~ 0 + treatment + block + (1 | species_mix)",
	method = "lmer"
)
```

```{r}
models_lmer |>
	dplyr::filter(warning != "")
```

The selection density effect model is giving us a sigular fit, but others are ok.

```{r}
models_lmer |>
	dplyr::filter(warning != "") |>
	pull(fit)
```

The model estimates the species_mix variance as exactly zero -- suggests that
the data don't support between-mixture random-intercept variation for this response.

Most of the `species_mix` levels have only 2 observations (4-species plots),
while one level has 32 observations (the 16-species plots).
We probably shouldn't have it as a random effect,
but I'm not sure it works as a fixed effect either since it's confounded with treatment. The `species_mix` terms might absorb most of the treatment effect.

The cleanest model is probably `reponse ~ 0 + treatment + block`.

We could do a separate exploratory model with species composition for only the 4-species plots, like `dens_selec ~ species_mix + block`, but will leave this for now.

## `lm` models

```{r}
models_lm <- fit_models(
	data = partitions,
	formula = "~ 0 + treatment + block",
	method = "lm"
)
```

Extract the coefficients and plot them

```{r}
results_lm <-
	models_lm |>
	mutate(
		tidy = purrr::map(
			fit,
			tidy,
			effects = "fixed",
			conf.int = TRUE
		)
	) |>
	unnest(tidy) |>
	mutate(
		term = fct_relevel(
			term,
			"treatment4-species",
			"treatment16-species",
			"blocksouth"
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 4.5

results_lm |>
	ggplot(aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
	geom_pointrange(shape = 21, fill = "white") +
	labs(x = "Term", y = "Estimate ± CI [95%]") +
	geom_hline(yintercept = 0, color = "blue") +
	coord_flip() +
	facet_wrap(~name, ncol = 1)
```

```{r}
models_lm <-
	models_lm |>
	mutate(
		resids = purrr::map(
			fit,
			DHARMa::simulateResiduals,
			plot = FALSE
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 9

purrr::walk2(
	models_lm$resids,
	models_lm$name,
	\(resids, name) plot(resids, title = name)
)
```

A fair ammount of deviation in the qq plots.
Could be due to long tails in the distribution of the response, see here:

```{r}
partitions |>
	ggplot(aes(x = net)) +
	geom_density() +

	partitions |>
		ggplot(aes(x = compl)) +
	geom_density() +

	partitions |>
		ggplot(aes(x = selec)) +
	geom_density() +

	partitions |>
		ggplot(aes(x = size_compl)) +
	geom_density() +

	partitions |>
		ggplot(aes(x = dens_compl)) +
	geom_density() +

	partitions |>
		ggplot(aes(x = size_selec)) +
	geom_density() +

	partitions |>
		ggplot(aes(x = dens_selec)) +
	geom_density()
```

## `glmmTMB` models

Let's try a t response distribution rather than Gaussian.

```{r}
models_tmb <- fit_models(
	data = partitions,
	formula = "~ 0 + treatment + block",
	method = "glmmTMB"
)
```

```{r}
models_tmb |>
	dplyr::filter(warning != "")
```

No warnings

```{r}
results_tmb <-
	models_tmb |>
	mutate(
		tidy = purrr::map(
			fit,
			tidy,
			effects = "fixed",
			conf.int = TRUE
		)
	) |>
	unnest(tidy) |>
	mutate(
		term = fct_relevel(
			term,
			"treatment4-species",
			"treatment16-species",
			"blocksouth"
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 9

results_tmb |>
	ggplot(aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
	geom_pointrange(shape = 21, fill = "white") +
	labs(x = "Term", y = "Estimate ± CI [95%]") +
	geom_hline(yintercept = 0, color = "blue") +
	coord_flip() +
	facet_wrap(~name, ncol = 1) +
	ggtitle("glmmTMB") +

	results_lm |>
		ggplot(aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
	geom_pointrange(shape = 21, fill = "white") +
	labs(x = "Term", y = "Estimate ± CI [95%]") +
	geom_hline(yintercept = 0, color = "blue") +
	coord_flip() +
	facet_wrap(~name, ncol = 1) +
	ggtitle("lm")
```

```{r}
models_tmb <-
	models_tmb |>
	mutate(
		resids = purrr::map(
			fit,
			DHARMa::simulateResiduals,
			plot = FALSE
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 9

purrr::walk2(
	models_tmb$resids,
	models_tmb$name,
	\(resids, name) plot(resids, title = name)
)
```

Residuals looking much better!

Printing output for all tmb models:

```{r}
models_tmb$fit
```
