---
title: "Basal area and density models"
---

```{r setup}
#| include: false

document_name <- tools::file_path_sans_ext(
	basename(knitr::current_input())
)

knitr::opts_chunk$set(
	fig.path = file.path("figures", document_name, "")
)

set.seed(20693)

source(here::here("code", "functions", "plotting.R"))
ggplot2::theme_set(theme_sbe())
```

I'm going to fit six models (as listed below) and create accompanying figures.

1. Including all plots:

- m1: `basal area ~ treatment + (1|block)`
- m2: `seedling density ~ treatment + (1|block)`

2. 16-species plots only:

- m3: `basal area ~ liana cutting + (1|block)`
- m4: `seedling density ~ liana cutting + (1|block)`

3. 4-species plots only:

- m5: `basal area ~ canopy complexity * generic richness + (1|block)`
- m6: `seedling density ~ canopy complexity * generic richness + (1|block)`

```{r}
library("tidyverse")
library("patchwork")
library("glmmTMB")
library("broom.mixed")
library("DHARMa")
library("ggdist")
library("emmeans")
```

```{r}
data_summed <-
	readRDS(here::here("data", "derived", "data_cleaned.rds")) |>
	filter(survival == 1) |> # only alive seedlings
	filter(census_no == '03') |>
	mutate(dbase_m = dbase_mm / 1000) |>
	mutate(basal_area = pi * (dbase_m / 2)^2) |>
	group_by(
		plot,
		block,
		species_mix,
		treatment,
		generic_diversity,
		struc_complexity
	) |>
	summarise(
		sum_basal_area = sum(basal_area, na.rm = TRUE),
		density = sum(survival, na.rm = TRUE),
		.groups = "drop"
	)
```

```{r}
data_summed |>
	ggplot(aes(x = sum_basal_area)) +
	geom_density() +
	data_summed |>
		ggplot(aes(x = density)) +
	geom_density()
```

Summed basal area is continuous and non-negative. If all values are positive and right-skewed, a Gamma model with a log link?

For seedling density, probably negative binomial

## 1. Including all plots

```{r}
data_summed |>
	filter(treatment != "16-species-cut") |>
	ggplot(aes(x = sum_basal_area, y = treatment, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +

	data_summed |>
		filter(treatment != "16-species-cut") |>
		ggplot(aes(x = density, y = treatment, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	plot_layout(guides = "collect")
```
```{r}
m1 <-
	glmmTMB(
		sum_basal_area ~ 0 + treatment + (1 | block),
		data = filter(data_summed, treatment != "16-species-cut"),
		family = Gamma(link = "log")
	)

m2 <-
	glmmTMB(
		density ~ 0 + treatment + (1 | block),
		data = filter(data_summed, treatment != "16-species-cut"),
		family = nbinom2(link = "log")
	)
```

Because we used a log link, we need to back transform estimates with `exp()`
to get them on the response scale.

```{r}
results_1 <-
	tibble(name = factor(c("Basal area", "Density")), fit = list(m1, m2)) |>
	mutate(
		tidy = purrr::map(
			fit,
			tidy,
			effects = "fixed",
			conf.int = TRUE,
			exponentiate = TRUE
		)
	)
```

```{r}
results_1 <-
	results_1 |>
	mutate(
		resids = purrr::map(
			fit,
			DHARMa::simulateResiduals,
			plot = FALSE
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 9

purrr::walk2(
	results_1$resids,
	results_1$name,
	\(resids, name) plot(resids, title = name)
)
```

```{r}
#| fig-height: 3
#| fig-width: 4.5

results_1 |>
	unnest(tidy) |>
	mutate(term = str_remove(term, "treatment")) |>
	mutate(
		term = fct_relevel(
			term,
			"1-species",
			"4-species",
			"16-species"
		),
		name = fct_relevel(
			name,
			"Basal area",
			"Density"
		)
	) |>
	ggplot(aes(
		x = term,
		y = estimate,
		ymin = conf.low,
		ymax = conf.high,
		colour = term
	)) +
	geom_pointrange(shape = 21, fill = "white") +
	labs(x = "Term", y = "Estimate ± CI [95%]") +
	coord_flip() +
	scale_colour_sbe() +
	facet_wrap(~name, ncol = 2, scales = "free_x")
```

Test if any difference between treatments:

```{r}
# basal area
emm_m1 <- emmeans(m1, ~treatment)
pairs(emm_m1, type = "response", infer = TRUE, level = 0.95)

# seedling density
emm_m2 <- emmeans(m2, ~treatment)
pairs(emm_m2, type = "response", infer = TRUE, level = 0.95)
```

## 2. 16-species plots only

```{r}
#| fig-height: 4
#| fig-width: 5

data_summed |>
	filter(species_mix == "16-species") |>
	ggplot(aes(x = sum_basal_area, y = treatment, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	data_summed |>
		filter(species_mix == "16-species") |>
		ggplot(aes(x = density, y = treatment, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	plot_layout(guides = "collect")
```

```{r}
m3 <-
	glmmTMB(
		sum_basal_area ~ 0 + treatment + (1 | block),
		data = filter(data_summed, species_mix == "16-species"),
		family = Gamma(link = "log")
	)

m4 <-
	glmmTMB(
		density ~ 0 + treatment + (1 | block),
		data = filter(data_summed, species_mix == "16-species"),
		family = nbinom2(link = "log")
	)
```

```{r}
results_2 <-
	tibble(name = factor(c("Basal area", "Density")), fit = list(m3, m4)) |>
	mutate(
		tidy = purrr::map(
			fit,
			tidy,
			effects = "fixed",
			conf.int = TRUE,
			exponentiate = TRUE
		)
	)
```

```{r}
results_2 <-
	results_2 |>
	mutate(
		resids = purrr::map(
			fit,
			DHARMa::simulateResiduals,
			plot = FALSE
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 9

purrr::walk2(
	results_2$resids,
	results_2$name,
	\(resids, name) plot(resids, title = name)
)
```

```{r}
#| fig-height: 3
#| fig-width: 4.5

results_2 |>
	unnest(tidy) |>
	mutate(term = str_remove(term, "treatment")) |>
	mutate(
		term = fct_relevel(
			term,
			"16-species",
			"16-species-cut"
		),
		name = fct_relevel(
			name,
			"Basal area",
			"Density"
		)
	) |>
	ggplot(aes(
		x = term,
		y = estimate,
		ymin = conf.low,
		ymax = conf.high,
		colour = term
	)) +
	geom_pointrange(shape = 21, fill = "white") +
	labs(x = "Term", y = "Estimate ± CI [95%]") +
	coord_flip() +
	scale_colour_sbe() +
	facet_wrap(~name, ncol = 2, scales = "free_x")
```

Test if any difference between treatments:

```{r}
# basal area
emm_m3 <- emmeans(m3, ~treatment)
pairs(emm_m3, type = "response", infer = TRUE, level = 0.95)

# seedling density
emm_m4 <- emmeans(m4, ~treatment)
pairs(emm_m4, type = "response", infer = TRUE, level = 0.95)
```

## 3. 4-species plots only:

Generic diversity is confounded with structural complexity,
and the full interaction model might be overparameterised beause one of the
four combinations is missing - there is no 4-genera, low-complexity combination.

So I don't think we can get clean, independent main effects.

We can try treating the 3 observed combinations as one 3-level factor.

```{r}
data_summed <- data_summed |>
	mutate(
		grichness_ccomplexity = interaction(
			generic_diversity,
			struc_complexity,
			sep = ", ",
			drop = TRUE
		)
	)
```

```{r}
#| fig-height: 4
#| fig-width: 5

data_summed |>
	filter(treatment == "4-species") |>
	ggplot(aes(
		x = sum_basal_area,
		y = grichness_ccomplexity,
		colour = treatment
	)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	data_summed |>
		filter(treatment == "4-species") |>
		ggplot(aes(x = density, y = grichness_ccomplexity, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	plot_layout(guides = "collect")
```

```{r}
m5 <-
	glmmTMB(
		sum_basal_area ~ 0 + grichness_ccomplexity + (1 | block),
		data = filter(data_summed, treatment == "4-species"),
		family = Gamma(link = "log")
	)

m6 <-
	glmmTMB(
		density ~ 0 + grichness_ccomplexity + (1 | block),
		data = filter(data_summed, treatment == "4-species"),
		family = nbinom2(link = "log")
	)
```

```{r}
results_3 <-
	tibble(name = factor(c("Basal area", "Density")), fit = list(m5, m6)) |>
	mutate(
		tidy = purrr::map(
			fit,
			tidy,
			effects = "fixed",
			conf.int = TRUE,
			exponentiate = TRUE
		)
	)
```

```{r}
results_3 <-
	results_3 |>
	mutate(
		resids = purrr::map(
			fit,
			DHARMa::simulateResiduals,
			plot = FALSE
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 9

purrr::walk2(
	results_3$resids,
	results_3$name,
	\(resids, name) plot(resids, title = name)
)
```

```{r}
#| fig-height: 3
#| fig-width: 4.5

results_3 |>
	unnest(tidy) |>
	mutate(
		name = fct_relevel(
			name,
			"Basal area",
			"Density"
		),
		treatment = "4-species"
	) |>
	ggplot(aes(
		x = term,
		y = estimate,
		ymin = conf.low,
		ymax = conf.high,
		colour = treatment
	)) +
	geom_pointrange(shape = 21, fill = "white") +
	labs(x = "Term", y = "Estimate ± CI [95%]") +
	coord_flip() +
	scale_colour_sbe() +
	facet_wrap(~name, ncol = 2, scales = "free_x")
```

Test if any difference between treatments:

```{r}
# basal area
emm_m3 <- emmeans(m5, ~grichness_ccomplexity)
pairs(emm_m3, type = "response", infer = TRUE, level = 0.95)

# seedling density
emm_m4 <- emmeans(m6, ~grichness_ccomplexity)
pairs(emm_m4, type = "response", infer = TRUE, level = 0.95)
```
