---
title: "Results report"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`'
format:
    gfm:
        toc: true
    typst:
        toc: true
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false
file_name <- knitr::current_input()

knitr::opts_chunk$set(
	fig.path = paste0(
		"figures/",
		sub("\\.rmarkdown$", "", basename(file_name)),
		"/",
		sep = ""
	)
)
set.seed(20654)

source(here::here("code", "functions", "plotting.R"))
ggplot2::theme_set(theme_sbe())
```

Boxplot figures of basal area and output data from the [`densize`](https://doi.org/10.1111/ele.14300) package.

```{r packages}
library("tidyverse")
library("patchwork")
library("here")
```

```{r data}
data <-
	readRDS(here::here("data", "derived", "data_cleaned.rds"))
```

# Summed basal diameter plots

```{r}
seedling_ba <-
	data |>
	filter(survival == 1) |> # only alive seedlings
	mutate(dbase_m = dbase_mm / 1000) |>
	mutate(basal_area = pi * (dbase_m / 2)^2) |>
	group_by(
		plot,
		census_yr,
		species_mix,
		treatment,
		generic_diversity,
		struc_complexity
	) |>
	summarise(sum_basal_area = sum(basal_area, na.rm = TRUE), .groups = "drop")
```

## Planting treatment

```{r}
seedling_ba |>
	ggplot(aes(
		x = census_yr,
		y = sum_basal_area,
		colour = treatment
	)) +
	geom_boxplot() +
	labs(y = "Summed basal area m2", x = "Census year") +
	scale_colour_sbe() +
	ggtitle("Summed basal area of planted seedlings per plot in 3rd full census")
```

Remember _n plots_ for liana cut plots is smaller than other treatments.
Calcluate per ha:

```{r}
seedling_ba |>
	filter(census_yr == "2023-2024") |>
	group_by(treatment) |>
	summarise(
		sum_basal_area = sum(sum_basal_area, na.rm = TRUE),
		n_plots = n_distinct(plot)
	) |>
	# 1 plot is 4ha
	mutate(basal_area_per_ha = (sum_basal_area / n_plots) / 4)
```

## Structural canopy complexity + generic diversity

```{r}
seedling_ba |>
	filter(census_yr == "2023-2024") |>
	filter(treatment == "4-species") |>
	ggplot(aes(
		x = struc_complexity,
		y = sum_basal_area,
		colour = treatment
	)) +
	geom_boxplot() +
	labs(y = "Summed basal area m2", x = "Structural canopy complexity") +
	scale_colour_sbe() +
	ggtitle(
		"Summed basal area of planted seedlings per plot in 3rd full census"
	) +
	seedling_ba |>
		filter(census_yr == "2023-2024") |>
		filter(treatment == "4-species") |>
		ggplot(aes(
			x = generic_diversity,
			y = sum_basal_area,
			colour = treatment
		)) +
	geom_boxplot() +
	labs(y = "Summed basal area m2", x = "Generic diversity") +
	scale_colour_sbe() +
	plot_layout(guides = "collect")
```

```{r}
seedling_ba |>
	filter(census_yr == "2023-2024") |>
	filter(treatment == "4-species") |>
	summarise(
		median_ba = median(sum_basal_area, na.rm = TRUE),
		q1 = quantile(sum_basal_area, 0.25, na.rm = TRUE),
		q3 = quantile(sum_basal_area, 0.75, na.rm = TRUE),
		.by = c(generic_diversity, struc_complexity)
	) |>
	mutate(tile_label = sprintf("%.1f\nIQR: %.1f–%.1f", median_ba, q1, q3)) |>
	ggplot(aes(
		x = generic_diversity,
		y = struc_complexity,
		fill = median_ba
	)) +
	geom_tile() +
	geom_text(
		aes(label = tile_label),
		size = 6,
		colour = "white"
	) +
	labs(fill = "Median basal area")
```

# Biodiversity effects

## Net biodiversity effects

```{r}
biodiv_data <-
	readRDS(here::here("data", "derived", "biodiv_effects.rds"))
```

```{r}
biodiv_data |>
	ggplot(aes(x = treatment, y = net, colour = treatment)) +
	geom_boxplot(outliers = FALSE) +
	geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
	geom_hline(yintercept = 0, linetype = 2) +
	scale_colour_sbe() +
	ggtitle("Net biodiversity effect")
```

## Selection and complementarity effects

```{r}
biodiv_data |>
	ggplot(aes(x = treatment, y = selec, colour = treatment)) +
	geom_boxplot(outliers = FALSE) +
	geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
	geom_hline(yintercept = 0, linetype = 2) +
	scale_colour_sbe() +
	ggtitle("Selection effect") +
	biodiv_data |>
		ggplot(aes(x = treatment, y = compl, colour = treatment)) +
	geom_boxplot(outliers = FALSE) +
	geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
	geom_hline(yintercept = 0, linetype = 2) +
	scale_colour_sbe() +
	ggtitle("Complementarity effect") +
	plot_layout(guides = "collect")
```

## Selection - size and density

```{r}
biodiv_data |>
	ggplot(aes(x = treatment, y = size_selec, colour = treatment)) +
	geom_boxplot(outliers = FALSE) +
	geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
	geom_hline(yintercept = 0, linetype = 2) +
	scale_colour_sbe() +
	ggtitle("Selection effect - size") +
	biodiv_data |>
		ggplot(aes(x = treatment, y = dens_selec, colour = treatment)) +
	geom_boxplot(outliers = FALSE) +
	geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
	geom_hline(yintercept = 0, linetype = 2) +
	scale_colour_sbe() +
	ggtitle("Selection effect - density") +
	plot_layout(guides = "collect")
```

## Complementarity - size and density

```{r}
biodiv_data |>
	ggplot(aes(x = treatment, y = size_compl, colour = treatment)) +
	geom_boxplot(outliers = FALSE) +
	geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
	geom_hline(yintercept = 0, linetype = 2) +
	scale_colour_sbe() +
	ggtitle("Complementarity effect - size") +
	biodiv_data |>
		ggplot(aes(x = treatment, y = dens_compl, colour = treatment)) +
	geom_boxplot(outliers = FALSE) +
	geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
	geom_hline(yintercept = 0, linetype = 2) +
	scale_colour_sbe() +
	ggtitle("Complementarity effect - density") +
	plot_layout(guides = "collect")
```
