---
title: "Additive partition of biodiversity effects"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: 
  gfm:
    toc: true
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

set.seed(20522)

ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
```

```{r}
#| output: FALSE

library("tidyverse")
library("here")
library("patchwork")
```

[Loreau & Hector 2001:](https://doi.org/10.1126/science.1064088)

> We measure the net biodiversity effect, $𝚫Y$, by the difference between the observed yield of a mixture and its expected yield under the null hypothesis that there is no selection effect or complementarity effect. This expected value is the weighted (by the initial relative abundance of species in mixture) average of the monoculture yields for the component species. Positive selection occurs if species with higher-than-average monoculture yields dominate the mixtures. The selection effect is measured by the covariance between the monoculture yield of species and their change in relative yield in the mixture. Finally, a positive complementarity effect occurs if species yields in a mixture are on average higher than expected on the basis of the weighted average monoculture yield of the component species. These various effects can be related by additive partition as follows.

$M_i =$ yield of species $i$ in monoculture

$Y_{O,i} =$ observed yield of species $i$ in the mixture

$Y_O = ∑_i Y_{O,i} =$ total observed yield of the mixture

$RY_{E,i} =$ expected relative yield of species $i$ in the mixture, which is simply its proportion seeded or planted

$RY_{O,i} = Y_{O,i}/M_i =$ observed relative yield of species $i$ in the mixture

$Y_{E,i} = RY_{E,i}M_i =$ expected yield of species $i$ in the mixture

$Y_E = ∑_iY_{E,i} =$ total expected yield of the mixture

$𝚫Y = Y_O - Y_E =$ deviation from total expected yield in the mixture

$𝚫RY_i = RY_{O,i} - RY_{E,i} =$ deviation from expected relative yield of species $i$ in the mixture

$N =$ number of species in the mixture

$$
𝚫Y = Y_O - Y_E = ∑_i RY_{O,i} M_i - ∑_i RY_{E,i} M_i = ∑_i 𝚫RY_iM_i = N\overline{𝚫RY} \overline{M} + N cov(𝚫RY,M)
$$

$N\overline{𝚫RY} \overline{M}$ is the complementarity effect and $N cov(𝚫RY,M)$ is the selection effect.

```{r}
data <- 
  readRDS(here::here("data", "derived", "data_cleaned.rds")) 
```

```{r}
data %>% 
  select(treatment, plot) %>% 
  group_by(treatment) %>% 
  summarise(n_distinct(plot))
```

Calculate basal area

```{r}
data <- 
  data %>% 
  mutate(dbase_m = dbase_mm / 1000) %>% 
  mutate(basal_area_m2 = pi * (dbase_m/2)^2)
```

Calculate yield (summed BA) at the species x plot level

```{r}
data_yield <- 
  data %>% 
  select(!survey_date:dbase_m & !census_no) %>% 
  pivot_wider(names_from = census_id,
              values_from = basal_area_m2) %>%
  # summed basal area  of each species in each treatment and plot
  group_by(genus_species, treatment, plot) %>%
  summarise(Y_Oi = sum(full_measurement_03, na.rm = TRUE),
            .groups = "drop") 
```

We now have $M_i$ (yield of species $i$ in monoculture) and $Y_{O,i}$ (observed yield of species $i$ in mixture). 
For $M_i$, I'm taking the mean across plots, so yield is at the plot level (4 ha).

```{r}
M_i <- 
  data_yield %>% 
  filter(treatment == "monoculture") %>% 
  group_by(genus_species) %>% 
  summarise(M_i = mean(Y_Oi))

Y_Oi <- 
  data_yield %>% 
  filter(treatment == "16-species" | treatment == "4-species") 
```

The total observed yield of the mixture $Y_O$ is the summed observed yield of all species in mixture $∑_i Y_{O,i}$ :

```{r}
Y_O <- 
  data_yield %>% 
  filter(treatment == "16-species" |
           treatment == "4-species") %>% 
  group_by(treatment, plot) %>% 
  summarise(Y_O = sum(Y_Oi),
            .groups = "drop")

glimpse(Y_O)
```

```{r}
Y_O %>% 
  ggplot(aes(x = treatment, y = Y_O)) +
  geom_boxplot() +
  geom_jitter() +
  ggtitle("Total observed yield")
```

We can calculate observed relative yield of species $i$ in the mixture 
as $RY_{O,i} = Y_{O,i}/M_i$ 
(i.e., the observed yield of species $i$ in the mixture divided by 
the yield of species $i$ in monoculture).

```{r}
RY_Oi <- 
  left_join(Y_Oi, M_i) %>% 
  # calculating separately for 16- and 4- species mix
  mutate(RY_Oi = Y_Oi / M_i)

glimpse(RY_Oi)
```
```{r}
RY_Oi %>% 
  ggplot(aes(y = genus_species, x = RY_Oi, 
             colour = treatment)) +
  geom_boxplot() +
  ggtitle("Observed relative yield")
```

The expected relative yield ($RY_{E,i}$) for each species in the 4-species mix is 0.25 and 0.0625 for each species in the 16-species mix (i.e., 1/4 and 1/16).

To calculate the expected yield of species $i$ in the mixture ($Y_{E,i}$) we can do $RY_{E,i}M_i$ (the expected relative yield multiplied by the yield of the species in monoculture):

```{r}
RY_Ei <- 
  data_yield %>% 
  filter(treatment == "16-species" |
           treatment == "4-species") %>% 
  select(genus_species, treatment, plot) %>% 
  left_join(M_i) %>% 
  mutate(exp_yield = case_when(
    treatment == "4-species" ~ M_i * 0.25,
    treatment == "16-species" ~ M_i * 0.0625)) 

glimpse(RY_Ei)
```

and the total expected yield of the mixture ($Y_E$) is those values summed across species $∑_iY_{E,i}$ :

```{r}
Y_E <- 
  RY_Oi %>% 
  left_join(RY_Ei) %>% 
  group_by(plot, treatment) %>% 
  summarise(total_exp_yield = sum(exp_yield),
            .groups = "drop")

glimpse(Y_E)
```

So now we can calculate $𝚫Y$, the deviation from total expected relative yield in the mixture as: $Y_{O} - Y_{E}$

```{r}
DY <- 
  full_join(Y_E, Y_O) %>% 
  mutate(DY = Y_O - total_exp_yield) 

glimpse(DY)
```

```{r}
DY %>% 
  ggplot(aes(x = treatment, y = DY)) +
  geom_boxplot() +
  geom_jitter() +
  ggtitle("The net biodiversity effect 
          (deviation from total expected yield in the mixture)")
```


and $𝚫RY_i$, the deviation from expected relative yield of species $i$ in the mixture as: $RY_{O,i} - RY_{E,i}$

```{r}
DRY_i <- 
  full_join(RY_Ei, RY_Oi) %>% 
  mutate(DRY_i = RY_Oi - exp_yield) 

glimpse(DRY_i)
```

```{r}
DRY_i %>% 
  ggplot(aes(y = genus_species, x = DRY_i, 
             colour = treatment)) +
  geom_boxplot() +
  ggtitle("Deviation from expected relative yield of species i in the mixture")
```

The complementarity effect is $N\overline{𝚫RY} \overline{M}$, the number of species in the mixture ($N$) multiplied by the mean $𝚫RY_i$ and mean $M_i$ across species.

```{r}
M_i_mean <-
  M_i %>% 
  summarise(mean(M_i)) %>% 
  pluck(1,1)

complementarity_effects <-
  DRY_i %>% 
  group_by(plot, treatment) %>% 
  summarise(mean_DRY_i = mean(DRY_i),
            .groups = "drop") %>% 
  mutate(complementarity_effect = mean_DRY_i * M_i_mean) %>% 
  select(plot, treatment, complementarity_effect)

glimpse(complementarity_effects)
```

and the selection effect $N cov(𝚫RY,M)$

```{r}

get_selection_effect <- function(plot, treatment, N, DRY_i_data, M_i_data) {
  DRY_i_data <- 
    DRY_i_data %>% 
    filter(treatment == !!treatment) %>% 
    filter(plot == !!plot) %>% 
    select(genus_species, DRY_i)
  
  M_i_data <- 
    M_i_data %>% 
    filter(genus_species %in% DRY_i_data$genus_species)
  
  selection_effect <- cov(x = DRY_i_data$DRY_i, y = M_i_data$M_i) * N
  
  tibble(
    plot = !!plot,
    treatment = !!treatment,
    selection_effect = selection_effect
  )
}

keys <- 
  DRY_i %>% 
  select(plot, treatment) %>% 
  distinct() %>% 
  mutate(N = as.numeric(str_extract(treatment, ".*(?=-)")))

selection_effects <- 
  purrr::pmap(
  .l = list(plot = keys$plot,
            treatment = keys$treatment,
            N = keys$N),
  .f = get_selection_effect,
  DRY_i_data = DRY_i,
  M_i_data = M_i) %>%
  list_rbind()

glimpse(selection_effects)
```


```{r}
complementarity_effects %>% 
  ggplot(aes(x = treatment,
             y = complementarity_effect)) +
  geom_jitter(width = 0.25) +
  
  selection_effects %>% 
  ggplot(aes(x = treatment,
             y = selection_effect)) +
  geom_jitter(width = 0.25)
```

```{r}
DY %>% 
  left_join(complementarity_effects) %>% 
  left_join(selection_effects) %>% glimpse
```

