---
title: "Partitioning density and size components of biodiversity effects"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: 
  gfm:
    toc: true
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

set.seed(20531)

ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
```

```{r}
library("tidyverse")
library("patchwork")
```

Attempting to get the the net, complementarity, and selection effects of biodiversity, 
using the [{densize} package](https://github.com/communityecologist/densize)
from [Tatsumi & Loreau 2023.](https://doi.org/10.1111/ele.14300)

```{r}
#devtools::install_github("communityecologist/densize")
library("densize")
```

> The specific data required are: (i) the initial density of each species sown, planted, or inoculated in each experimental unit (e.g. plot, chemostat) (`init.dens`); (ii) the biomass y of each species in each unit (`final.yield`); and (iii) the plant density d of each species in each unit at the time of biomass measurement (`final.dens`). Note that the biomass y does not have to be recorded at the individual level but only at the species level. Plant density d can be measured either by counting the individuals directly or by dividing the species biomass y by the mean per-plant biomass of representative individuals in each sampling unit.

`densize(init.dens, final.dens, final.yield)`

`init.dens` : A matrix or data frame consisting of initial plant density (the number of plants or seeds sown per area).

`final.dens` : A matrix or data frame consisting of final plant density (the number of plants survived per area).

`final.yield` : A matrix or data frame consisting of final yield (yield per area).

## Create `init.dens`

```{r}
data <- 
  readRDS(here::here("data", "derived", "data_cleaned.rds")) 
```

```{r}
init_dens <- 
  data %>% 
  filter(census_id == "full_measurement_01") %>% 
  group_by(plot, genus_species) %>% 
  summarise(n = n_distinct(plant_id), .groups = "drop") %>% 
  pivot_wider(id_cols = plot,
              names_from = genus_species,
              values_from = n) %>% 
  mutate(across(everything(), \(x) replace_na(x, 0))) %>% 
  arrange(plot) %>% 
  column_to_rownames(var = "plot")
```

```{r}
glimpse(init_dens)
```

## Create `final.dens`

```{r}
final_dens <- 
  data %>% 
  filter(census_id == "full_measurement_03") %>% 
  filter(survival == 1) %>% 
  group_by(plot, genus_species) %>% 
  summarise(n = n_distinct(plant_id), .groups = "drop") %>% 
  pivot_wider(id_cols = plot,
              names_from = genus_species,
              values_from = n) %>% 
  mutate(across(everything(), \(x) replace_na(x, 0))) %>% 
  arrange(plot) %>% 
  column_to_rownames(var = "plot")
```

```{r}
glimpse(final_dens)
```

## Create `final.yield`

```{r}
final_yield <- 
  data %>% 
  filter(census_id == "full_measurement_03") %>% 
  filter(survival == 1) %>% 
  mutate(dbase_m = dbase_mm / 1000) %>% 
  mutate(basal_area_m2 = pi * (dbase_m/2)^2) %>% 
  group_by(plot, genus_species) %>% 
  summarise(sum = sum(basal_area_m2, na.rm = TRUE), .groups = "drop") %>% 
  pivot_wider(id_cols = plot,
              names_from = genus_species,
              values_from = sum) %>% 
  mutate(across(everything(), \(x) replace_na(x, 0))) %>% 
  arrange(plot) %>% 
  column_to_rownames(var = "plot")
```

```{r}
glimpse(final_yield)
```

## Run function

```{r}
densize_out <- 
  densize(init.dens = init_dens, 
        final.dens = final_dens, 
        final.yield = final_yield) %>% 
  as_tibble()

metadata <- 
  data %>% 
  select(plot, treatment) %>% 
  distinct() %>% 
  filter(treatment != "monoculture") %>% 
  arrange(plot)

result <- 
  bind_cols(densize_out, metadata) %>% 
  mutate(compl = dens.compl + size.compl,
         selec = dens.selec + size.selec,
         net = compl + selec) %>% 
  mutate(treatment = fct_relevel(treatment, 
                                 "4-species", 
                                 "16-species",
                                 "16-species-cut"))

glimpse(result)
```

## Plotting

```{r}
result %>% 
  ggplot(aes(x = treatment, y = net)) +
  geom_boxplot(outliers = FALSE) +
  geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
  ggtitle("Net biodiversity effect") +
  
  result %>% 
  ggplot(aes(x = treatment, y = selec)) +
  geom_boxplot(outliers = FALSE) +
  geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
  ggtitle("Selection effect") +
  
  result %>% 
  ggplot(aes(x = treatment, y = compl)) +
  geom_boxplot(outliers = FALSE) +
  geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
  ggtitle("Complementarity effect")
```

Looks like:

- Net biodiversity effect might increase with Liana cutting
- Selection effect might decrease with species richness
- Complementarity effect might increase with species richness

```{r}
result %>% 
  ggplot(aes(x = treatment, y = dens.selec)) +
  geom_boxplot(outliers = FALSE) +
  geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
  ggtitle("Selection effect - density") +
  
  result %>% 
  ggplot(aes(x = treatment, y = size.selec)) +
  geom_boxplot(outliers = FALSE) +
  geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
  ggtitle("Selection effect - size") 
```

Differences in selection effect are mostly driven by density rather than size?

```{r}
result %>% 
  ggplot(aes(x = treatment, y = dens.compl)) +
  geom_boxplot(outliers = FALSE) +
  geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
  ggtitle("Complementarity effect - density") +
  
  result %>% 
  ggplot(aes(x = treatment, y = size.compl)) +
  geom_boxplot(outliers = FALSE) +
  geom_jitter(width = 0.25, shape = 16, alpha = 0.5) +
  ggtitle("Complementarity effect - size") 
```


```{r}
dens_size <-
  data %>% 
  filter(survival == 1) %>% 
  mutate(dbase_m = dbase_mm / 1000) %>% 
  mutate(basal_area_m2 = pi * (dbase_m/2)^2) %>% 
  filter(census_id == "full_measurement_03") %>% 
  group_by(plot, treatment) %>% 
  summarise(dens = n_distinct(plant_id),
            size = sum(basal_area_m2)) %>% 
  mutate(size = size / dens) 

dens_size %>% 
  ggplot(aes(x = dens, y = size, colour = treatment)) +
  geom_point() +
  geom_text(data = 
              filter(dens_size, size > 0.015 | dens > 300),
            aes(label = plot),
            position = position_nudge(y = -0.001) ) +
  labs(y = "Plant size (mean basal area per plant m2)",
       x = "Plant density (number per plot)",
       title = "Density–size relationships")
```

No clear density-size pattern!

Taking a look at which species are included in those outlier plots:

```{r}
data %>%
  filter(plot == "114" |
           plot == "065" |
           plot == "001" |
           plot == "002" |
           plot == "004") %>% 
  select(plot, treatment, genus_species) %>% 
  distinct()
  
```


# Biodiversity effects over time

Above we are looking from the 1st census to the 3rd.
We can also break it down to look between the 1st and 2nd censuses 
and the 2nd and 3rd censuses.

```{r}
# Census 1 to 2
init_dens_01 <- 
  data %>% 
  filter(census_id == "full_measurement_01") %>% 
  group_by(plot, genus_species) %>% 
  summarise(n = n_distinct(plant_id), .groups = "drop") %>% 
  pivot_wider(id_cols = plot,
              names_from = genus_species,
              values_from = n) %>% 
  mutate(across(everything(), \(x) replace_na(x, 0))) %>% 
  arrange(plot) %>% 
  column_to_rownames(var = "plot")

final_dens_02 <- 
  data %>% 
  filter(census_id == "full_measurement_02") %>% 
  filter(survival == 1) %>% 
  group_by(plot, genus_species) %>% 
  summarise(n = n_distinct(plant_id), .groups = "drop") %>% 
  pivot_wider(id_cols = plot,
              names_from = genus_species,
              values_from = n) %>% 
  mutate(across(everything(), \(x) replace_na(x, 0))) %>% 
  arrange(plot) %>% 
  column_to_rownames(var = "plot")

final_yield_02 <- 
  data %>% 
  filter(census_id == "full_measurement_02") %>% 
  filter(survival == 1) %>% 
  mutate(dbase_m = dbase_mm / 1000) %>% 
  mutate(basal_area_m2 = pi * (dbase_m/2)^2) %>% 
  group_by(plot, genus_species) %>% 
  summarise(sum = sum(basal_area_m2, na.rm = TRUE), .groups = "drop") %>% 
  pivot_wider(id_cols = plot,
              names_from = genus_species,
              values_from = sum) %>% 
  mutate(across(everything(), \(x) replace_na(x, 0))) %>% 
  arrange(plot) %>% 
  column_to_rownames(var = "plot")

densize_out_1_2 <- 
  densize(init.dens = init_dens_01, 
        final.dens = final_dens_02, 
        final.yield = final_yield_02) %>% 
  as_tibble()

result_1_2 <- 
  bind_cols(densize_out_1_2, metadata) %>% 
  mutate(compl = dens.compl + size.compl,
         selec = dens.selec + size.selec,
         net = compl + selec) %>% 
  mutate(treatment = fct_relevel(treatment, 
                                 "4-species", 
                                 "16-species",
                                 "16-species-cut"))

# Census 2 to 3

init_dens_02 <- 
  data %>% 
  filter(census_id == "full_measurement_02") %>% 
  group_by(plot, genus_species) %>% 
  summarise(n = n_distinct(plant_id), .groups = "drop") %>% 
  pivot_wider(id_cols = plot,
              names_from = genus_species,
              values_from = n) %>% 
  mutate(across(everything(), \(x) replace_na(x, 0))) %>% 
  arrange(plot) %>% 
  column_to_rownames(var = "plot")

final_dens_03 <- 
  data %>% 
  filter(census_id == "full_measurement_03") %>% 
  filter(survival == 1) %>% 
  group_by(plot, genus_species) %>% 
  summarise(n = n_distinct(plant_id), .groups = "drop") %>% 
  pivot_wider(id_cols = plot,
              names_from = genus_species,
              values_from = n) %>% 
  mutate(across(everything(), \(x) replace_na(x, 0))) %>% 
  arrange(plot) %>% 
  column_to_rownames(var = "plot")

final_yield_03 <- 
  data %>% 
  filter(census_id == "full_measurement_03") %>% 
  filter(survival == 1) %>% 
  mutate(dbase_m = dbase_mm / 1000) %>% 
  mutate(basal_area_m2 = pi * (dbase_m/2)^2) %>% 
  group_by(plot, genus_species) %>% 
  summarise(sum = sum(basal_area_m2, na.rm = TRUE), .groups = "drop") %>% 
  pivot_wider(id_cols = plot,
              names_from = genus_species,
              values_from = sum) %>% 
  mutate(across(everything(), \(x) replace_na(x, 0))) %>% 
  arrange(plot) %>% 
  column_to_rownames(var = "plot")

densize_out_2_3 <- 
  densize(init.dens = init_dens_02, 
        final.dens = final_dens_03, 
        final.yield = final_yield_03) %>% 
  as_tibble()

result_2_3 <- 
  bind_cols(densize_out_2_3, metadata) %>% 
  mutate(compl = dens.compl + size.compl,
         selec = dens.selec + size.selec,
         net = compl + selec) %>% 
  mutate(treatment = fct_relevel(treatment, 
                                 "4-species", 
                                 "16-species",
                                 "16-species-cut"))
```

```{r}
result_cens <- 
  bind_rows(list("02" = result_1_2, "03" = result_2_3),
          .id = "census") %>% glimpse
```

```{r}
delta <- 
  result_cens %>% 
  select(census, treatment, census, net, plot) %>%
  pivot_wider(names_from = census, values_from = net, id_cols = c(treatment, plot)) %>%
  mutate(delta = `03` - `02`) %>% 
  select(treatment, plot, delta)

result_cens %>% 
  pivot_longer(cols = where(is.numeric)) %>%
  filter(name == "net") %>% 
  left_join(delta) %>% 
  ggplot(aes(x = census, y = value, colour = delta)) +
  geom_line(aes(group = plot, colour = delta)) +
  geom_point() +
  scale_colour_gradientn(colours = colorspace::divergingx_hcl(n = 20, palette = "RdBu"),
                           limits = c(-1, 1)) +
  labs(y = "basal area m2 per plot") +
  facet_wrap(~treatment) +
  ggtitle("Net biodiversity effect")
```

```{r}
result_cens %>% 
  pivot_longer(cols = where(is.numeric)) %>%
  filter(name == "compl" | name == "selec" ) %>% 
  ggplot(aes(x = census, y = value, colour = name)) +
  geom_line(aes(group = interaction(plot, name), colour = name),
              linewidth = 0.25,
              alpha = 0.4) +
  geom_point() +
  labs(y = "basal area m2 per plot") +
  facet_wrap(~treatment)
```

Selection and complementarity effects increase in magnitude over time.

```{r}
result_cens %>% 
  pivot_longer(cols = where(is.numeric)) %>%
  filter(name == "size.selec" | name == "dens.compl" ) %>% 
  ggplot(aes(x = census, y = value, colour = name)) +
  geom_line(aes(group = interaction(plot, name), colour = name),
              linewidth = 0.25,
              alpha = 0.4) +
  geom_point() +
  labs(y = "basal area m2 per plot",
       title = "Selection effect") +
  facet_wrap(~treatment) +
  
  result_cens %>% 
  pivot_longer(cols = where(is.numeric)) %>%
  filter(name == "size.compl" |name == "dens.compl") %>% 
  ggplot(aes(x = census, y = value, colour = name)) +
  geom_line(aes(group = interaction(plot, name), colour = name),
              linewidth = 0.25,
              alpha = 0.4) +
  geom_point() +
  labs(y = "basal area m2 per plot",
       title = "Complementarity effect") +
  facet_wrap(~treatment) +
  
  plot_layout(ncol = 1)
  
```
