---
title: "Basal area and density models"
---

```{r setup}
#| include: false

document_name <- tools::file_path_sans_ext(
	basename(knitr::current_input())
)

knitr::opts_chunk$set(
	fig.path = file.path("figures", document_name, "")
)

set.seed(20693)

source(here::here("code", "functions", "plotting.R"))
ggplot2::theme_set(theme_sbe())
```

I'm going to fit six models (as listed below) and create accompanying figures.

1. Including all plots:

- m1: `basal area ~ treatment + (1|block)`
- m2: `seedling density ~ treatment + (1|block)`

2. 16-species plots only:

- m3: `basal area ~ liana cutting + (1|block)`
- m4: `seedling density ~ liana cutting + (1|block)`

3. 4-species plots only:

- m5: `basal area ~ canopy complexity * generic richness + (1|block)`
- m6: `seedling density ~ canopy complexity * generic richness + (1|block)`

```{r}
library("tidyverse")
library("patchwork")
library("glmmTMB")
library("broom.mixed")
library("DHARMa")
library("ggdist")
library("emmeans")
```

```{r}
data_summed <-
	readRDS(here::here("data", "derived", "data_cleaned.rds")) |>
	filter(survival == 1) |> # only alive seedlings
	filter(census_no == '03') |>
	mutate(dbase_m = dbase_mm / 1000) |>
	mutate(basal_area = pi * (dbase_m / 2)^2) |>
	group_by(
		plot,
		block,
		species_mix,
		treatment,
		generic_richness,
		struc_complexity
	) |>
	summarise(
		sum_basal_area = sum(basal_area, na.rm = TRUE),
		density = sum(survival, na.rm = TRUE),
		.groups = "drop"
	)
```

```{r}
#| fig-height: 3
#| fig-width: 5
data_summed |>
	ggplot(aes(x = sum_basal_area)) +
	geom_density() +
	data_summed |>
		ggplot(aes(x = density)) +
	geom_density()
```

Summed basal area is continuous and non-negative with a right skew.
Lognormal distibution is usually used for e.g. size, mass (exponential growth).
The alternative way to do this would be to log the data and fit a gaussian.

For seedling density, we probably want to model a negative binomial response
distribution (discrete, non-negative, right skew).

## 1. Including all plots

```{r}
data_summed |>
	filter(treatment != "16-species-cut") |>
	group_by(treatment) |>
	summarise(n = n_distinct(plot))
```

```{r}
data_summed |>
	filter(treatment != "16-species-cut") |>
	ggplot(aes(x = sum_basal_area, y = treatment, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +

	data_summed |>
		filter(treatment != "16-species-cut") |>
		ggplot(aes(x = density, y = treatment, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	plot_layout(guides = "collect")
```

For density, I'm modelling the dispersion parameter as a function of treatment.
The single-species plots have a more variance than the other two groups (see figure above).
When I ran the model treating the dispersion parameter as a constant,
it "violated the assumption of a homogeneous dispersion parameter across groups" - DHARMa.

```{r}
m1 <-
	glmmTMB(
		sum_basal_area ~ 0 + treatment + (1 | block),
		data = filter(data_summed, treatment != "16-species-cut"),
		family = lognormal(link = "log")
	)

m2 <-
	glmmTMB(
		density ~ 0 + treatment + (1 | block),
		dispformula = ~treatment,
		data = filter(data_summed, treatment != "16-species-cut"),
		family = nbinom2(link = "log")
	)
```

Because we used a log link, we need to back transform estimates with `exp()`
to get them on the response scale.

Beacuse we use a log-link in all our models, contrasts become ratios after back-transformation, so:

- ratio of `1` = groups have the same estimated response
- `1.25` = first group is 25% higher than the second
- `0.80` = first group is 20% lower than the second
- `2` = first group has twice the estimated response
- `0.50` = first group has half the estimated response

For confidence intervals:

- If the entire ratio CI is above `1`, the first group is estimated to have the higher response
- If the entire CI is below `1`, the first group is estimated to have the lower response
- If the CI includes `1`, the results are compatible with no difference, as well as the range of effects covered by the interval

```{r}
results_1 <-
	tibble(name = factor(c("Basal area", "Density")), fit = list(m1, m2)) |>
	mutate(
		tidy = purrr::map(
			fit,
			tidy,
			effects = "fixed",
			conf.int = TRUE,
			exponentiate = TRUE
		)
	)
```

```{r}
results_1 <-
	results_1 |>
	mutate(
		resids = purrr::map(
			fit,
			DHARMa::simulateResiduals,
			plot = FALSE
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 9

purrr::walk2(
	results_1$resids,
	results_1$name,
	\(resids, name) plot(resids, title = name)
)
```

```{r}
#| fig-height: 3
#| fig-width: 4.5

results_1 |>
	unnest(tidy) |>
	mutate(term = str_remove(term, "treatment")) |>
	mutate(
		term = fct_relevel(
			term,
			"1-species",
			"4-species",
			"16-species"
		),
		name = fct_relevel(
			name,
			"Basal area",
			"Density"
		)
	) |>
	ggplot(aes(
		x = term,
		y = estimate,
		ymin = conf.low,
		ymax = conf.high,
		colour = term
	)) +
	geom_pointrange(shape = 21, fill = "white") +
	labs(x = "Term", y = "Estimate ± CI [95%]") +
	coord_flip() +
	scale_colour_sbe() +
	facet_wrap(~name, ncol = 2, scales = "free_x")
```

### Results

Test if any difference between treatments:

```{r}
#| fig-height: 3
#| fig-width: 4.5
# basal area
emm_m1 <- emmeans(m1, ~treatment)

pairs(
	emm_m1,
	type = "response",
	side = "two-sided",
	infer = TRUE,
	level = 0.95,
	adjust = "none"
)

plot(
	emm_m1,
	type = "response",
	side = "two-sided",
	level = 0.95,
	comparisons = TRUE,
	adjust = "none"
) +
	ggtitle("Basal area") +
	theme_sbe()
```

- Basal area
    - Basal area was estimated to be similar in 1- and 4-species plots (ratio: 1.03; 95% CI: 0.79–1.34). The interval was compatible with basal area being approximately 21% lower to 34% higher in 1-species plots
    - Basal area in 1-species plots was estimated to be 0.79 times that in 16-species plots (95% CI: 0.61–1.02), corresponding to an estimated 21% lower basal area. The interval ranged from approximately 39% lower to 2% higher and included 1
    - Basal area in 4-species plots was estimated to be 0.77 times that in 16-species plots (95% CI: 0.60–0.99), corresponding to an estimated 23% lower basal area. The entire interval was below 1, indicating that basal area was estimated to be approximately 1–40% lower in 4-species plots

```{r}
#| fig-height: 3
#| fig-width: 4.5
# seedling density
emm_m2 <- emmeans(m2, ~treatment)

pairs(
	emm_m2,
	type = "response",
	side = "two-sided",
	infer = TRUE,
	level = 0.95,
	adjust = "none"
)

plot(
	emm_m2,
	type = "response",
	side = "two-sided",
	level = 0.95,
	comparisons = TRUE,
	adjust = "none"
) +
	ggtitle("Seedling density") +
	theme_sbe()
```

- Seedling density
    - Seedling density was estimated to be similar in 1- and 4-species plots (ratio: 1.02; 95% CI: 0.76–1.38)
    - Seedling density in 1-species plots was estimated to be 1.13 times that in 16-species plots (95% CI: 0.86–1.50), corresponding to an estimated 13% higher density, with the interval ranging from approximately 14% lower to 50% higher
    - Seedling density in 4-species plots was estimated to be 1.11 times that in 16-species plots (95% CI: 0.93–1.32), corresponding to an estimated 11% higher density, with the interval ranging from approximately 7% lower to 32% higher

### Modelling richness as continuous

Because the richness levels increase multiplicatively,
define a `richness_step` variable that gives equally spaced values 0, 1, 2. With log-link models, the coefficient then represents the multiplicative change in the response for each fourfold increase in species richness.

```{r}
data_summed <-
	data_summed |>
	mutate(
		species_richness = readr::parse_number(as.character(treatment)),
		richness_step = log(species_richness, base = 4)
	)
```

```{r}
m1_cont <-
	glmmTMB(
		sum_basal_area ~ richness_step + (1 | block),
		data = filter(data_summed, treatment != "16-species-cut"),
		family = lognormal(link = "log")
	)

m2_cont <-
	glmmTMB(
		density ~ richness_step + (1 | block),
		dispformula = ~treatment,
		data = filter(data_summed, treatment != "16-species-cut"),
		family = nbinom2(link = "log")
	)
```

Compare to `m1` and `m2`

```{r}
anova(m1_cont, m1)
anova(m2_cont, m2)
```

Continuous models have lower AIC... but also lower DF

```{r}
results_1_cont <-
	tibble(
		name = factor(c("Basal area", "Density")),
		fit = list(m1_cont, m2_cont)
	) |>
	mutate(
		tidy = purrr::map(
			fit,
			tidy,
			effects = "fixed",
			conf.int = TRUE,
			exponentiate = TRUE
		)
	)
```

```{r}
results_1_cont <-
	results_1_cont |>
	mutate(
		resids = purrr::map(
			fit,
			DHARMa::simulateResiduals,
			plot = FALSE
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 9

purrr::walk2(
	results_1_cont$resids,
	results_1_cont$name,
	\(resids, name) plot(resids, title = name)
)
```

Inspect the slope on the response scale:

```{r}
#| fig-height: 3
#| fig-width: 4.5

results_1_cont |>
	unnest(tidy) |>
	filter(term != "(Intercept)") |>
	mutate(
		name = fct_relevel(
			name,
			"Basal area",
			"Density"
		)
	) |>
	ggplot(aes(
		x = name,
		y = estimate,
		ymin = conf.low,
		ymax = conf.high
	)) +
	geom_pointrange(shape = 21, fill = "white") +
	labs(
		x = "Term",
		y = "Response ratio per fourfold increase in species richness ± CI [95%]"
	) +
	geom_hline(
		yintercept = 1,
		linetype = "dashed"
	) +
	scale_y_log10() +
	coord_flip() +
	theme_sbe()
```

A point to the right of 1 indicates a positive slope and
to the left - a negative slope.

Get model predictions on the link scale and exponentiate them:

```{r}
get_predictions <- \(model, response_name) {
	emmeans(
		model,
		~richness_step,
		at = list(richness_step = 0:2)
	) |>
		broom::tidy(conf.int = TRUE) |>
		transmute(
			response = response_name,
			richness_step,
			species_richness = 4^richness_step,
			estimate = exp(estimate),
			conf.low = exp(conf.low),
			conf.high = exp(conf.high)
		)
}

predictions <-
	bind_rows(
		get_predictions(m1_cont, "Basal area"),
		get_predictions(m2_cont, "Seedling density")
	)
```

```{r}
#| fig-height: 3
#| fig-width: 5
predictions |>
	ggplot(
		aes(
			x = richness_step,
			y = estimate,
			ymin = conf.low,
			ymax = conf.high
		)
	) +
	geom_ribbon(
		aes(group = response),
		alpha = 0.15
	) +
	geom_line() +
	geom_point(size = 3) +
	facet_wrap(
		vars(response),
		scales = "free_y"
	) +
	scale_x_continuous(
		breaks = 0:2,
		labels = c("1", "4", "16")
	) +
	labs(
		x = "Species richness",
		y = "Predicted response"
	) +
	theme_sbe()
```

### Results (continuous)

- Basal area
  - Basal area was estimated to _increase_ with species richness
  - Each fourfold increase in species richness—from 1 to 4 species or from 4 to 16 species—was associated with an estimated 1.14-fold increase in basal area (95% CI: 0.99–1.30), corresponding to an estimated increase of approximately 14%
  - The interval ranged from approximately 1% lower to 30% higher and included a response ratio of 1, so the data remained compatible with no change
- Seedling density
  - Seedling density was estimated to _decrease_ with species richness
  - Each fourfold increase in species richness was associated with an estimated density ratio of 0.93 (95% CI: 0.82–1.04), corresponding to an estimated decrease of approximately 7%
  - The interval ranged from approximately 18% lower to 4% higher and included a response ratio of 1, so the data remained compatible with no change

## 2. 16-species plots only

```{r}
data_summed |>
	filter(species_mix == "16-species") |>
	group_by(treatment) |>
	summarise(n = n_distinct(plot))
```

```{r}
#| fig-height: 3
#| fig-width: 5

data_summed |>
	filter(species_mix == "16-species") |>
	ggplot(aes(x = sum_basal_area, y = treatment, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	data_summed |>
		filter(species_mix == "16-species") |>
		ggplot(aes(x = density, y = treatment, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	plot_layout(guides = "collect")
```

```{r}
m3 <-
	glmmTMB(
		sum_basal_area ~ 0 + treatment + (1 | block),
		data = filter(data_summed, species_mix == "16-species"),
		family = lognormal(link = "log")
	)

m4 <-
	glmmTMB(
		density ~ 0 + treatment + (1 | block),
		data = filter(data_summed, species_mix == "16-species"),
		family = nbinom2(link = "log")
	)
```

```{r}
results_2 <-
	tibble(name = factor(c("Basal area", "Density")), fit = list(m3, m4)) |>
	mutate(
		tidy = purrr::map(
			fit,
			tidy,
			effects = "fixed",
			conf.int = TRUE,
			exponentiate = TRUE
		)
	)
```

```{r}
results_2 <-
	results_2 |>
	mutate(
		resids = purrr::map(
			fit,
			DHARMa::simulateResiduals,
			plot = FALSE
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 9

purrr::walk2(
	results_2$resids,
	results_2$name,
	\(resids, name) plot(resids, title = name)
)
```

```{r}
#| fig-height: 3
#| fig-width: 4.5

results_2 |>
	unnest(tidy) |>
	mutate(term = str_remove(term, "treatment")) |>
	mutate(
		term = fct_relevel(
			term,
			"16-species",
			"16-species-cut"
		),
		name = fct_relevel(
			name,
			"Basal area",
			"Density"
		)
	) |>
	ggplot(aes(
		x = term,
		y = estimate,
		ymin = conf.low,
		ymax = conf.high,
		colour = term
	)) +
	geom_pointrange(shape = 21, fill = "white") +
	labs(x = "Term", y = "Estimate ± CI [95%]") +
	coord_flip() +
	scale_colour_sbe() +
	facet_wrap(~name, ncol = 2, scales = "free_x")
```

### Results

```{r}
#| fig-height: 3
#| fig-width: 4.5
# basal area
emm_m3 <- emmeans(m3, ~treatment)

pairs(
	emm_m3,
	type = "response",
	side = "two-sided",
	infer = TRUE,
	level = 0.95,
	reverse = TRUE,
	adjust = "none"
)

plot(
	emm_m3,
	type = "response",
	side = "two-sided",
	level = 0.95,
	reverse = TRUE,
	comparisons = TRUE,
	adjust = "none"
) +
	ggtitle("Basal area") +
	theme_sbe()
```

- Basal area
    - Basal area in liana-cut plots was estimated to be 1.25 times that in uncut 16-species plots (95% CI: 0.97–1.61), corresponding to an estimated 25% increase
    - The confidence interval ranged from approximately 3% lower to 61% higher and included a ratio of 1. The data therefore remain compatible with no difference, although the point estimate suggests higher basal area following liana cutting

```{r}
#| fig-height: 3
#| fig-width: 4.5
# seedling density
emm_m4 <- emmeans(m4, ~treatment)

pairs(
	emm_m4,
	type = "response",
	side = "two-sided",
	infer = TRUE,
	level = 0.95,
	reverse = TRUE,
	adjust = "none"
)

plot(
	emm_m4,
	type = "response",
	side = "two-sided",
	level = 0.95,
	reverse = TRUE,
	comparisons = TRUE,
	adjust = "none"
) +
	ggtitle("Seedling density") +
	theme_sbe()
```


- Seedling density
    - Seedling density in liana-cut plots was estimated to be 1.26 times that in uncut 16-species plots (95% CI: 1.05–1.50), corresponding to an estimated 26% increase
    - The entire confidence interval was above 1, indicating that density was estimated to be approximately 5–50% higher in liana-cut plots
    - The results support higher seedling density following liana cutting, but the confidence interval around the magnitude of that increase is large

## 3. 4-species plots only:

```{r}
data_summed |>
	filter(treatment == "4-species") |>
	group_by(generic_richness, struc_complexity) |>
	summarise(n = n_distinct(plot))
```


```{r}
#| fig-height: 4
#| fig-width: 5

data_summed |>
	filter(treatment == "4-species") |>
	ggplot(aes(x = sum_basal_area, y = struc_complexity, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	facet_grid(vars(generic_richness)) +
	data_summed |>
		filter(treatment == "4-species") |>
		ggplot(aes(x = density, y = struc_complexity, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	plot_layout(guides = "collect") +
	facet_grid(rows = vars(generic_richness))
```

```{r}
m5 <-
	glmmTMB(
		sum_basal_area ~ struc_complexity * generic_richness + (1 | block),
		data = filter(data_summed, treatment == "4-species"),
		family = lognormal(link = "log")
	)

m6 <-
	glmmTMB(
		density ~ struc_complexity * generic_richness + (1 | block),
		data = filter(data_summed, treatment == "4-species"),
		family = nbinom2(link = "log")
	)
```

```{r}
results_3 <-
	tibble(name = factor(c("Basal area", "Density")), fit = list(m5, m6)) |>
	mutate(
		tidy = purrr::map(
			fit,
			tidy,
			effects = "fixed",
			conf.int = TRUE,
			exponentiate = TRUE
		)
	)
```

```{r}
results_3 <-
	results_3 |>
	mutate(
		resids = purrr::map(
			fit,
			DHARMa::simulateResiduals,
			plot = FALSE
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 9

purrr::walk2(
	results_3$resids,
	results_3$name,
	\(resids, name) plot(resids, title = name)
)
```

Borderline DHARMa test for the basal area model `m5`,
but I think this is due to the outlier in the 4-genera, high complexity group rather than true heteroscedasticity.

### Results

```{r}
#| fig-height: 3
#| fig-width: 4.5
# basal area
emm_m5 <- emmeans(m5, ~ struc_complexity * generic_richness)

pairs(
	emm_m5,
	type = "response",
	side = "two-sided",
	infer = TRUE,
	level = 0.95,
	reverse = TRUE,
	adjust = "none"
)

plot(
	emm_m5,
	type = "response",
	side = "two-sided",
	level = 0.95,
	reverse = TRUE,
	comparisons = TRUE,
	adjust = "none"
) +
	ggtitle("Basal area") +
	theme_sbe()
```

- Basal area
    - At two genera, basal area under high canopy complexity was estimated to be 1.22 times that under low complexity (95% CI: 0.75–2.00). The interval included 1 and remained compatible with no difference
    - At four genera, basal area under high canopy complexity was estimated to be 1.69 times that under low complexity (95% CI: 1.01–2.84), indicating an estimated increase of approximately 1–184%
    - The high-versus-low complexity ratio was estimated to be 1.38 times greater at four genera than at two genera (interaction ratio: 1.38; 95% CI: 0.68–2.83). This interval included 1, so the magnitude of the interaction remains uncertain
    - Comparing generic richness within each complexity level, basal area in four-genera plots was estimated to be 0.87 times that in two-genera plots under low complexity (95% CI: 0.51–1.48) and 1.20 times that in two-genera plots under high complexity (95% CI: 0.74–1.96). Both intervals included 1

```{r}
#| fig-height: 3
#| fig-width: 4.5
# basal area
emm_m6 <- emmeans(m6, ~ struc_complexity * generic_richness)

pairs(
	emm_m6,
	type = "response",
	side = "two-sided",
	infer = TRUE,
	level = 0.95,
	reverse = TRUE,
	adjust = "none"
)

plot(
	emm_m6,
	type = "response",
	side = "two-sided",
	level = 0.95,
	reverse = TRUE,
	comparisons = TRUE,
	adjust = "none"
) +
	ggtitle("Seedling density") +
	theme_sbe()
```

- Seedling density
    - At two genera, seedling density under high canopy complexity was estimated to be 1.46 times that under low complexity (95% CI: 1.05–2.04), indicating an estimated increase of approximately 5–104%
    - At four genera, seedling density under high canopy complexity was estimated to be 1.63 times that under low complexity (95% CI: 1.17–2.26), indicating an estimated increase of approximately 17–126%
    - The high-versus-low complexity ratio was estimated to be 1.11 times greater at four genera than at two genera (interaction ratio: 1.11; 95% CI: 0.70–1.77). This interval included 1, providing no clear evidence that the canopy-complexity effect differed with generic richness
    - Comparing generic richness within each complexity level, density in four-genera plots was estimated to be 1.10 times that in two-genera plots under low complexity (95% CI: 0.79–1.54) and 1.23 times that in two-genera plots under high complexity (95% CI: 0.88–1.70). Both intervals included 1

## Results summary

Across the species richness treatments (all plots) (not the continuous models), basal area was estimated to be 23% lower in 4-species plots than in 16-species plots (95% CI: 1–40% lower).
The other basal area comparison intervals and all seedling density comparison intervals included 1 and remained compatible with no difference.

In the liana-cutting comparisons (16-species plots), the basal area interval included 1. However, seedling density was estimated to be 26% higher in liana-cut plots than in uncut plots, with the 95% confidence interval indicating an increase of approximately 5–50%.

Within 4-species plots, high canopy complexity was associated with higher basal area at four genera and higher seedling density at both two and four genera.
However, the interaction intervals included 1 for both responses, so there was no clear evidence that the effect of canopy complexity differed between the two generic-richness levels.
The generic-richness contrasts within each canopy-complexity level also remained compatible with no difference.
