---
title: "_Dipterocarpus conformis_ check"
toc: false
---

```{r setup}
#| include: false

document_name <- tools::file_path_sans_ext(
	basename(knitr::current_input())
)

knitr::opts_chunk$set(
	fig.path = file.path("figures", document_name, "")
)

set.seed(as.numeric(Sys.Date()))

source(here::here("code", "functions", "plotting.R"))
ggplot2::theme_set(theme_sbe())
```

>"_Dipterocarpus conformis_ seedlings were not available for replanting, meaning this one species was not planted. Do we know: were the dead individuals of that species left unreplaced or replaced with seedlings of the other species?"

```{r packages}
library("tidyverse")
library("patchwork")
library("here")
```

```{r data}
data <-
	readRDS(here("data", "derived", "data_cleaned.rds"))
```


```{r}
data |>
	filter(genus_species == "Dipterocarpus_conformis") |>
	group_by(cohort) |>
	summarise(n_distinct(plant_id))
```

_Dipterocarpus conformis_ seedlings exist only in cohort 1.

```{r}
# find D. conformis seedlings which were dead in 1st census
dead_dc <-
	data |>
	filter(genus_species == "Dipterocarpus_conformis") |>
	filter(census_no == "01") |>
	filter(survival == "0") |>
	distinct(plant_id, plot, line, position)
```

```{r}
# find any cohort 2 seedlings in the same position
# as the dead D. conformis seedlings

empty_positions <-
	dead_dc |>
	select(plot, line, position)

data |>
	filter(census_no != "01", cohort == 2) |>
	inner_join(empty_positions) |>
	distinct(genus_species)
```

Hmm, it looks like other species were planted in place of the dead *D. conformis.*

But surely not in the *D. conformis* 1-species mixes?

```{r}
data |>
	filter(census_no != "01", cohort == 2) |>
	inner_join(empty_positions) |>
	distinct(treatment)
```

No, just the other mixes.

Did the replanting exactly replace seedlings which had died?

```{r}
data |>
	group_by(cohort) |>
	summarise(n_distinct(plant_id))
```

There are fewer seedlings in cohort 2 than cohort 1.

```{r}
n_dead <-
	data |>
	filter(census_no == "01") |>
	filter(survival == 0) |>
	summarise(n_distinct(plant_id)) |>
	pluck(1, 1)
```

`{r} n_dead` cohort 1 seedlings died in the first census - fewer than the total number of cohort 2 seedlings.
I initially thought that this meant not all seedlings were replanted,
however, some of the replanted seedlings might have died between planting
(2009-2010) and their first census (the second full census of the SBE, 2011-2013).

Some figures comparing numbers of dead cohort 1 seedlings in the first census
with numbers of cohort 2 seedlings in the second census:

```{r}
# find seedlings (of any sp) which were dead in the 1st census
dead_cen1 <-
	data |>
	filter(census_no == "01") |>
	filter(survival == 0) |>
	distinct(plant_id, plot, genus_species, treatment) |>
	group_by(plot, genus_species, treatment) |>
	summarise(n_dead = n_distinct(plant_id))

# find replanted seedlings in cohort 2
replanted <-
	data |>
	filter(cohort == 2) |>
	distinct(plant_id, plot, genus_species, treatment) |>
	group_by(plot, genus_species, treatment) |>
	summarise(n_replanted = n_distinct(plant_id))

dead_v_replant <-
	full_join(dead_cen1, replanted) |>
	mutate(
		n_dead = replace_na(n_dead, 0),
		n_replanted = replace_na(n_replanted, 0)
	) |>
	mutate(n_dead = -n_dead) |>
	pivot_longer(cols = c(n_dead, n_replanted))
```

```{r}
#| fig-width: 10
#| fig-height: 10

dead_v_replant |>
	filter(treatment == "1-species") |>
	ggplot(aes(x = plot, y = value, fill = name)) +
	geom_col() +
	facet_wrap(~genus_species, scales = "free_x") +
	geom_hline(yintercept = 0, linetype = 2) +
	ggtitle("Dead vs replanted - 1-species plots") +
	theme_sbe(base_size = 15)
```

```{r}
#| fig-width: 10
#| fig-height: 10

dead_v_replant |>
	filter(treatment == "4-species") |>
	ggplot(aes(x = plot, y = value, fill = name)) +
	geom_col() +
	facet_wrap(~genus_species, scales = "free_x") +
	geom_hline(yintercept = 0, linetype = 2) +
	ggtitle("Dead vs replanted - 4-species plots") +
	theme_sbe(base_size = 15) +
	theme(axis.text.x = element_text(angle = 90))
```

```{r}
#| fig-width: 10
#| fig-height: 10

dead_v_replant |>
	filter(treatment == "16-species" | treatment == "16-species-cut") |>
	ggplot(aes(x = plot, y = value, fill = name)) +
	geom_col() +
	facet_wrap(~genus_species, scales = "free_x") +
	geom_hline(yintercept = 0, linetype = 2) +
	ggtitle("Dead vs replanted - 16-species plots") +
	theme_sbe(base_size = 15) +
	theme(axis.text.x = element_text(angle = 90))
```
