---
title: "Basal area and density models"
---

```{r setup}
#| include: false

document_name <- tools::file_path_sans_ext(
	basename(knitr::current_input())
)

knitr::opts_chunk$set(
	fig.path = file.path("figures", document_name, "")
)

set.seed(20693)

source(here::here("code", "functions", "plotting.R"))
ggplot2::theme_set(theme_sbe())
```

I'm going to fit six models (as listed below) and create accompanying figures.

1. Including all plots:

- m1: `basal area ~ treatment + (1|block)`
- m2: `seedling density ~ treatment + (1|block)`

2. 16-species plots only:

- m3: `basal area ~ liana cutting + (1|block)`
- m4: `seedling density ~ liana cutting + (1|block)`

3. 4-species plots only:

- m5: `basal area ~ canopy complexity * generic richness + (1|block)`
- m6: `seedling density ~ canopy complexity * generic richness + (1|block)`

```{r}
library("tidyverse")
library("patchwork")
library("glmmTMB")
library("broom.mixed")
library("DHARMa")
library("ggdist")
library("emmeans")
```

```{r}
data_summed <-
	readRDS(here::here("data", "derived", "data_cleaned.rds")) |>
	filter(survival == 1) |> # only alive seedlings
	filter(census_no == '03') |>
	mutate(dbase_m = dbase_mm / 1000) |>
	mutate(basal_area = pi * (dbase_m / 2)^2) |>
	group_by(
		plot,
		block,
		species_mix,
		treatment,
		generic_diversity,
		struc_complexity
	) |>
	summarise(
		sum_basal_area = sum(basal_area, na.rm = TRUE),
		density = sum(survival, na.rm = TRUE),
		.groups = "drop"
	)
```

```{r}
data_summed |>
	ggplot(aes(x = sum_basal_area)) +
	geom_density() +
	data_summed |>
		ggplot(aes(x = density)) +
	geom_density()
```

Summed basal area is continuous and non-negative with a right skew.
Lognormal distibution is usually used for e.g. size, mass (exponential growth).
The alternative way to do this would be to log the data and fit a gaussian.

For seedling density, we probably want to model a negative binomial response
distribution (discrete, non-negative, right skew).

## 1. Including all plots

```{r}
data_summed |>
	filter(treatment != "16-species-cut") |>
	group_by(treatment) |>
	summarise(n = n_distinct(plot))
```

```{r}
data_summed |>
	filter(treatment != "16-species-cut") |>
	ggplot(aes(x = sum_basal_area, y = treatment, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +

	data_summed |>
		filter(treatment != "16-species-cut") |>
		ggplot(aes(x = density, y = treatment, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	plot_layout(guides = "collect")
```

For density, I'm modelling the dispersion parameter as a function of treatment.
The single-species plots have a more variance than the other two groups (see figure above).
When I ran the model treating the dispersion parameter as a constant,
it "violated the assumption of a homogeneous dispersion parameter across groups" - DHARMa.

```{r}
m1 <-
	glmmTMB(
		sum_basal_area ~ 0 + treatment + (1 | block),
		data = filter(data_summed, treatment != "16-species-cut"),
		family = lognormal(link = "log")
	)

m2 <-
	glmmTMB(
		density ~ 0 + treatment + (1 | block),
		dispformula = ~treatment,
		data = filter(data_summed, treatment != "16-species-cut"),
		family = nbinom2(link = "log")
	)
```

Because we used a log link, we need to back transform estimates with `exp()`
to get them on the response scale.

```{r}
results_1 <-
	tibble(name = factor(c("Basal area", "Density")), fit = list(m1, m2)) |>
	mutate(
		tidy = purrr::map(
			fit,
			tidy,
			effects = "fixed",
			conf.int = TRUE,
			exponentiate = TRUE
		)
	)
```

```{r}
results_1 <-
	results_1 |>
	mutate(
		resids = purrr::map(
			fit,
			DHARMa::simulateResiduals,
			plot = FALSE
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 9

purrr::walk2(
	results_1$resids,
	results_1$name,
	\(resids, name) plot(resids, title = name)
)
```

```{r}
#| fig-height: 3
#| fig-width: 4.5

results_1 |>
	unnest(tidy) |>
	mutate(term = str_remove(term, "treatment")) |>
	mutate(
		term = fct_relevel(
			term,
			"1-species",
			"4-species",
			"16-species"
		),
		name = fct_relevel(
			name,
			"Basal area",
			"Density"
		)
	) |>
	ggplot(aes(
		x = term,
		y = estimate,
		ymin = conf.low,
		ymax = conf.high,
		colour = term
	)) +
	geom_pointrange(shape = 21, fill = "white") +
	labs(x = "Term", y = "Estimate ± CI [95%]") +
	coord_flip() +
	scale_colour_sbe() +
	facet_wrap(~name, ncol = 2, scales = "free_x")
```

Test if any difference between treatments:

```{r}
# basal area
emm_m1 <- emmeans(m1, ~treatment)
pairs(
	emm_m1,
	type = "response",
	side = "two-sided",
	infer = TRUE,
	level = 0.95,
	adjust = "tukey"
)

# seedling density
emm_m2 <- emmeans(m2, ~treatment)
pairs(
	emm_m2,
	type = "response",
	side = "two-sided",
	infer = TRUE,
	level = 0.95,
	adjust = "tukey"
)
```

Beacuse we use a log-link in all our models, contrasts become ratios after back-transformation, so:

- ratio of `1` = groups have the same estimated response
- `1.25` = first group is 25% higher than the second
- `0.80` = first group is 20% lower than the second
- `2` = first group has twice the estimated response
- `0.50` = first group has half the estimated response

For confidence intervals:

- If the entire ratio CI is above `1`, the first group is estimated to have the higher response
- If the entire CI is below `1`, the first group is estimated to have the lower response
- If the CI includes `1`, the results are compatible with no difference, as well as the range of effects covered by the interval

### Results

- Basal area
    - Basal area was estimated to be similar in 1- and 4-species plots (ratio: 1.03; Tukey-adjusted 95% CI: 0.75–1.42). The interval was compatible with basal area being approximately 25% lower to 42% higher in 1-species plots
    - Basal area in 1-species plots was estimated to be 0.79 times that in 16-species plots (95% CI: 0.58–1.07), corresponding to an estimated 21% lower basal area. However, the interval ranged from approximately 42% lower to 7% higher
    - Basal area in 4-species plots was estimated to be 0.77 times that in 16-species plots (95% CI: 0.57–1.03), corresponding to an estimated 23% lower basal area. The interval ranged from approximately 43% lower to 3% higher
- Seedling density
    - Seedling density was estimated to be similar in 1- and 4-species plots (ratio: 1.02; Tukey-adjusted 95% CI: 0.72–1.46)
    - Seedling density in 1-species plots was estimated to be 1.13 times that in 16-species plots (95% CI: 0.81–1.59), corresponding to an estimated 13% higher density, with the interval ranging from approximately 19% lower to 59% higher
    - Seedling density in 4-species plots was estimated to be 1.11 times that in 16-species plots (95% CI: 0.90–1.36), corresponding to an estimated 11% higher density, with the interval ranging from approximately 10% lower to 36% higher
- All intervals include a ratio of one, suggesting that there is no clear difference in the estimated response between groups

## 2. 16-species plots only

```{r}
data_summed |>
	filter(species_mix == "16-species") |>
	group_by(treatment) |>
	summarise(n = n_distinct(plot))
```

```{r}
#| fig-height: 4
#| fig-width: 5

data_summed |>
	filter(species_mix == "16-species") |>
	ggplot(aes(x = sum_basal_area, y = treatment, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	data_summed |>
		filter(species_mix == "16-species") |>
		ggplot(aes(x = density, y = treatment, colour = treatment)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	plot_layout(guides = "collect")
```

```{r}
m3 <-
	glmmTMB(
		sum_basal_area ~ 0 + treatment + (1 | block),
		data = filter(data_summed, species_mix == "16-species"),
		family = lognormal(link = "log")
	)

m4 <-
	glmmTMB(
		density ~ 0 + treatment + (1 | block),
		data = filter(data_summed, species_mix == "16-species"),
		family = nbinom2(link = "log")
	)
```

```{r}
results_2 <-
	tibble(name = factor(c("Basal area", "Density")), fit = list(m3, m4)) |>
	mutate(
		tidy = purrr::map(
			fit,
			tidy,
			effects = "fixed",
			conf.int = TRUE,
			exponentiate = TRUE
		)
	)
```

```{r}
results_2 <-
	results_2 |>
	mutate(
		resids = purrr::map(
			fit,
			DHARMa::simulateResiduals,
			plot = FALSE
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 9

purrr::walk2(
	results_2$resids,
	results_2$name,
	\(resids, name) plot(resids, title = name)
)
```

```{r}
#| fig-height: 3
#| fig-width: 4.5

results_2 |>
	unnest(tidy) |>
	mutate(term = str_remove(term, "treatment")) |>
	mutate(
		term = fct_relevel(
			term,
			"16-species",
			"16-species-cut"
		),
		name = fct_relevel(
			name,
			"Basal area",
			"Density"
		)
	) |>
	ggplot(aes(
		x = term,
		y = estimate,
		ymin = conf.low,
		ymax = conf.high,
		colour = term
	)) +
	geom_pointrange(shape = 21, fill = "white") +
	labs(x = "Term", y = "Estimate ± CI [95%]") +
	coord_flip() +
	scale_colour_sbe() +
	facet_wrap(~name, ncol = 2, scales = "free_x")
```

Test if any difference between treatments:

```{r}
# basal area
emm_m3 <- emmeans(m3, ~treatment)
pairs(
	emm_m3,
	type = "response",
	side = "two-sided",
	infer = TRUE,
	level = 0.95,
	reverse = TRUE,
	adjust = "none" # only 2 groups so no adjustment needed
)

# seedling density
emm_m4 <- emmeans(m4, ~treatment)
pairs(
	emm_m4,
	type = "response",
	side = "two-sided",
	infer = TRUE,
	level = 0.95,
	reverse = TRUE,
	adjust = "none"
)
```

### Results

- Basal area
    - Basal area in liana-cut plots was estimated to be 1.25 times that in uncut 16-species plots (95% CI: 0.97–1.61), corresponding to an estimated 25% increase
    - The confidence interval ranged from approximately 3% lower to 61% higher and included a ratio of 1. The data therefore remain compatible with no difference, although the point estimate suggests higher basal area following liana cutting
- Seedling density
    - Seedling density in liana-cut plots was estimated to be 1.26 times that in uncut 16-species plots (95% CI: 1.05–1.50), corresponding to an estimated 26% increase
    - The entire confidence interval was above 1, indicating that density was estimated to be approximately 5–50% higher in liana-cut plots
    - the results support higher seedling density following liana cutting, but the confidence interval around the magnitude of that increase is large

Even though the confidence intervals for the two seedling density estimates overlap (see figure above), this does not directly measure uncertainty in their difference.
The model contrast incorporates the covariance between the estimates and supports higher seedling density in liana-cut plots.

## 3. 4-species plots only:

Generic diversity is confounded with structural complexity,
and the full interaction model might be overparameterised beause one of the
four combinations is missing - there is no 4-genera, low-complexity combination.

So I don't think we can get clean, independent main effects.

We can try treating the 3 observed combinations as one 3-level factor.

```{r}
data_summed <- data_summed |>
	mutate(
		grichness_ccomplexity = interaction(
			generic_diversity,
			struc_complexity,
			sep = ", ",
			drop = TRUE
		)
	)
```

```{r}
data_summed |>
	filter(treatment == "4-species") |>
	group_by(grichness_ccomplexity) |>
	summarise(n = n_distinct(plot))
```

```{r}
#| fig-height: 4
#| fig-width: 5
data_summed |>
	filter(treatment == "4-species" & generic_diversity == "2-genera") |>
	ggplot(aes(
		x = sum_basal_area,
		y = struc_complexity,
		colour = treatment
	)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	data_summed |>
		filter(treatment == "4-species" & generic_diversity == "2-genera") |>
		ggplot(aes(
			x = density,
			y = struc_complexity,
			colour = treatment
		)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	plot_layout(guides = "collect") +
	plot_annotation(title = "Generic richness held at two")
```

```{r}
#| fig-height: 4
#| fig-width: 5
data_summed |>
	filter(treatment == "4-species" & struc_complexity == "high") |>
	ggplot(aes(
		x = sum_basal_area,
		y = generic_diversity,
		colour = treatment
	)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	data_summed |>
		filter(treatment == "4-species" & struc_complexity == "high") |>
		ggplot(aes(
			x = density,
			y = generic_diversity,
			colour = treatment
		)) +
	geom_swarm(shape = 16) +
	scale_colour_sbe() +
	plot_layout(guides = "collect") +
	plot_annotation(title = "Canopy structural complexity held at `high`")
```

```{r}
m5 <-
	glmmTMB(
		sum_basal_area ~ 0 + grichness_ccomplexity + (1 | block),
		data = filter(data_summed, treatment == "4-species"),
		family = lognormal(link = "log")
	)

m6 <-
	glmmTMB(
		density ~ 0 + grichness_ccomplexity + (1 | block),
		data = filter(data_summed, treatment == "4-species"),
		family = nbinom2(link = "log")
	)
```

```{r}
results_3 <-
	tibble(name = factor(c("Basal area", "Density")), fit = list(m5, m6)) |>
	mutate(
		tidy = purrr::map(
			fit,
			tidy,
			effects = "fixed",
			conf.int = TRUE,
			exponentiate = TRUE
		)
	)
```

```{r}
results_3 <-
	results_3 |>
	mutate(
		resids = purrr::map(
			fit,
			DHARMa::simulateResiduals,
			plot = FALSE
		)
	)
```

```{r}
#| fig-height: 6
#| fig-width: 9

purrr::walk2(
	results_3$resids,
	results_3$name,
	\(resids, name) plot(resids, title = name)
)
```

```{r}
#| fig-height: 3
#| fig-width: 4.5
results_3 |>
	unnest(tidy) |>
	mutate(
		name = fct_relevel(
			name,
			"Basal area",
			"Density"
		),
		treatment = "4-species"
	) |>
	ggplot(aes(
		x = term,
		y = estimate,
		ymin = conf.low,
		ymax = conf.high,
		colour = treatment
	)) +
	geom_pointrange(shape = 21, fill = "white") +
	labs(x = "Term", y = "Estimate ± CI [95%]") +
	coord_flip() +
	scale_colour_sbe() +
	facet_wrap(~name, ncol = 2, scales = "free_x")
```

Test if any difference between treatments:

```{r}
# basal area
emm_m5 <- emmeans(m5, ~grichness_ccomplexity)
contrast(
	emm_m5,
	method = list(
		"canopy complexity low vs high (at 2 genera)" = c(1, -1, 0),
		"generic richness 2 vs 4 (at high complexity)" = c(0, 1, -1)
	),
	type = "response",
	side = "two-sided",
	infer = TRUE,
	level = 0.95,
	adjust = "holm"
)

# seedling density
emm_m6 <- emmeans(m6, ~grichness_ccomplexity)
contrast(
	emm_m6,
	method = list(
		"canopy complexity low vs high (at 2 genera)" = c(1, -1, 0),
		"generic richness 2 vs 4 (at high complexity)" = c(0, 1, -1)
	),
	type = "response",
	side = "two-sided",
	infer = TRUE,
	level = 0.95,
	adjust = "holm"
)
```

### Results

- Basal area
    - Basal area under low canopy complexity was 0.82 times (0.47-1.43) that under high complexity when generic richness was held at 2 genera
    - Basal area under generic richness 2 was 1.16 times (0.71-1.90) that in the 4 genera treatment when canopy complexity was held at `high`
- Seedling density
    - Seedling density under low canopy complexity was 0.68 times (0.45-1.04) that under high complexity when generic richness was held at 2 genera
    - When canopy complexity was held at `high` seedling density was nearly identical between the two- and four- genera treatments (ratio 1.01, 0.70-1.46)
- All intervals include one, suggesting that there is no clear difference in the estimated response between groups.

## Results summary

Across the species richness treatments (all plots), all 95% confidence intervals for the basal area and seedling density ratios included 1,
meaning the estimates remained compatible with no difference.

The confidence intervals for comparisons among the observed generic richness and canopy complexity combinations (4-species plots) likewise included 1.

In the liana cutting comparisons (16-species plots), the basal area ratio also showed no difference between groups. However, seedling density was estimated to be 26% higher in liana-cut plots than in uncut plots, with the 95% confidence interval indicating an increase of approximately 5% to 50%.
